---
title: "3D relationships between RDI, RCI and D in TTWAs"
author: "Jon Minton"
date: "15 February 2017"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
# Code chunk example from http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
library(knitr)
library(rgl)
opts_knit$set(root.dir=normalizePath('../'))
knit_hooks$set(webgl = hook_webgl)

```

# Aim

This document will (hopefully) allow D, RCI and RDI to be explored interactively for each of the TTWAs. 
For each TTWA, the first period is indicated with a blue sphere, the last with a red sphere, and intermediate spheres along the path are grey. 


```{r setup2, include = F, echo = F}

require(pacman)

pacman::p_load(
  tidyverse,
  readr, readxl,
  forcats,
  stringr,  
  purrr,
  rgeos,
  ggplot2,
  tmap
)

source("scripts/from_gavin/RCI.R")


D <- function(minority, total){
  majority      <- total - minority
  MINORITY      <- sum(minority)
  MAJORITY      <- sum(majority)
  
  p1 <- minority / MINORITY
  p2 <- majority / MAJORITY 
  
  ad <- abs(p1 - p2)
  
  out <- 0.5 * sum(ad)
  out
}



# imd
dist_to_centre <- read_csv("data/lsoa_2011_by_dist_to_centres.csv")
dta <- read_csv("data/imd/imd_id_tidied.csv", col_types = "icdd")

#SIMD
dist_to_centre_scot <- read_csv("data/dz_2011_by_dist_to_centres.csv")
dta_scot <- read_csv("data/simd/simd_combined_on_2011.csv")

# Inputs to RCI are 
#povvec
#popvec
#order


# AS the inputs to the RCI function simply include a vector giving order by distance, 
# by changing the order vector to density (highest to lowest) a relative densification 
# index (not quite the same as RCO) can be calculated using the same function. 
# The degree of similarity of dissimilarity between RCI and RDI can be taken as an indicator 
# of polycentricity (maybe)


dta %>% 
  inner_join(dist_to_centre) %>%
  mutate(pdens = pop_total / area) %>% 
  group_by(place, year) %>% 
  nest() %>% 
  mutate(
    pvec = map(data, ~ .[["pop_id"]]),
    tvec = map(data, ~ .[["pop_total"]]),
    ordr = map(data, ~ order(.[["distance"]])),
    ordr_dens = map(data, ~ order(.[["pdens"]], decreasing = T))
         ) %>% 
  mutate(rci_val = pmap_dbl(list(pvec, tvec, ordr), RCI)) %>% 
  mutate(rdi_val = pmap_dbl(list(pvec, tvec, ordr_dens), RCI)) %>% 
  mutate(d_val = pmap_dbl(list(pvec, tvec), D)) %>% 
  select(place, year, rci = rci_val, rdi = rdi_val, d = d_val) %>% 
  ungroup -> rci_by_year_place


# Same but for Scotland
dta_scot %>% 
  inner_join(dist_to_centre_scot, by = c("dz_2011" = "dz")) %>%
  mutate(pdens = pop_total / area) %>%
  select(dz_2011, place, year, pop_id = pop_incomedeprived, pop_total, distance, pdens) %>% 
  group_by(place, year) %>% 
  nest() %>% 
  mutate(
    pvec = map(data, ~ .[["pop_id"]]),
    tvec = map(data, ~ .[["pop_total"]]),
    ordr = map(data, ~ order(.[["distance"]])),
    ordr_dens = map(data, ~ order(.[["pdens"]], decreasing = T))
  ) %>% 
  mutate(rci_val = pmap_dbl(list(pvec, tvec, ordr), RCI)) %>% 
  mutate(rdi_val = pmap_dbl(list(pvec, tvec, ordr_dens), RCI)) %>% 
  mutate(d_val = pmap_dbl(list(pvec, tvec), D)) %>% 
  select(place, year, rci = rci_val, rdi = rdi_val, d = d_val) %>% 
  ungroup -> rci_by_year_place_scot


# Combine England w/ Scotland 

rci_by_year_place <- bind_rows(
  rci_by_year_place, 
  rci_by_year_place_scot
) %>% 
  filter(place != "Cardiff")
```

## Including Plots

You can also embed plots, for example:

```{r 3d_function, echo=FALSE, cache = T}

show_3d_change <- function(PLACE){
  rci_by_year_place %>% 
    filter(place == PLACE) %>% 
    arrange(year) -> tmp
  
  n <- nrow(tmp)
  

  with(tmp, rgl::plot3d(rdi, rci, d, type = "l", expand = 1.10))

  with(tmp, rgl::plot3d(rdi, rci, d, type = "s", 
                        col = c("red", rep("black", n - 2), "blue"), 
                        alpha = c(1, rep(0.5, n-2), 1), add = T))
  rgl::planes3d(1,0,0, 0, alpha = 0.2, add = T)
  rgl::planes3d(0,1,0, 0, alpha = 0.2, add = T)
  title3d(PLACE)
  NULL
}
```

```{r do_glasgow, webgl = T}
show_3d_change("Glasgow")


```
show_3d_change("London")
show_3d_change("Edinburgh")
show_3d_change("Leicester")
show_3d_change("Cambridge")
show_3d_change("Leicester")

